A programming model for temporal coordination (in music)

### 2014-05-01 paper/paper.tex
generated .tex file

The underlying programming model of SonicPi provides a way to separate
the ordering of effects from the timing of
effects. Figure~\ref{three-chord-example} shows an example program
where three chords are played in sequence, combining simple notions of 
parallel, timed, and ordered effects. The first three statements play
the notes of a C major chord in parallel. A \sleepOp{} statement then 
provides a \emph{barrier}

Thus ``$\sleep{} t$'' communicates that, after it has been evaluated, at least 
$t$ seconds has elapsed since the last \sleepOp{}. This provides a minimum
time. In between calls to \sleepOp{}, any other statements can (with some limits)
be considered task parallel.

@@

\begin{itemize}
\item A computational model of virtual and real time 
      structured by an \emph{applicative functor} abstraction.

\item We show that the applicative approach extends
      to a monadic approach, which can be composed with additional
      monads to capture other useful notions of effect in \lang{} programs,
      such as random numbers (Section~\ref{}).
\end{itemize}

Previously reported on the language~\cite{aaron2013sonic}

@@

Figure~\ref{sleep-examples} shows four similar programs which each
have different internal behaviours for \sleepOp, illustrating the
semantics of \sleepOp{}. The first three take 3s to execute and the
last takes 4s to execute, with the behaviours:
%
\begin{enumerate}[(a)]
\item{3s -- sleeps for 1s then sleeps for 2s (two sleeps performed);}
\item{3s -- performs a computation lasting 1s, ignores
the first \sleepOp{} since its minimum deadline has been reached, 
and then sleeps for 2s (one sleep performed);}
\item{3s -- performs a computation lasting 2s, which means that
the first \sleepOp{} is ignored, and the second \sleepOp{} waits
for only 1s to reach its minimum deadline (half a sleep performed);}
\item{4s -- performs a computation lasting 2s, thus 
the first \sleepOp{} is ignored, then performs a computation lasting
2s, thus the second \sleepOp{} is ignored (no sleeps performed).}
\end{enumerate}

@@

Since \sleepOp{} is the only operation that changes the virtual
time, we can easily specify the definition of $\vtime{-}$ over all programs:
%
\begin{definition}
Virtual time is specified for statements of \lang{} programs 
by the following (ordered) cases:
%
\begin{align*}
\vtime{\sleep t} & = t \\ 
\vtime{P; Q} & = \vtime{P} + \vtime{Q} \\
\vtime{-} & = 0
\end{align*}
%
\ie{}, the virtual time is $0$ 
for any statment other than \sleepOp{} or sequential composition.
\label{sleep-spec}
\end{definition}

Providing exact deadlines in real-time system is difficult due 
to noise and overhead caused by execution. We do not wish to sweep
this under the carpet in \lang{} or in our models. We therefore define
a relation $\approx$ on actual time (non virtual times), where:
%%
\begin{equation}
s \approx t
\equiv
t - \varepsilon \leq s \leq t + \varepsilon 
\end{equation}
%
for some value of $\varepsilon$ which is the maximum negligible 
time value with respect to the application at hand. In the case of
\lang{}, this is equal to the scheduling delay for the SuperCollider
engine (and is machine dependent), which is roughly \note{X}.

@@

The virtual time and actual time of a single sleep statement (in
isolation) are roughly the same, \ie{}, $\vtime{\sleep t} \approx
\etime{\sleep t}$ and thus $\etime{\sleep t} \approx t$ (by the
specification in Definition~\ref{sleep-spec}). Note that this holds
only when \sleepOp{} is used in isolation, that is, when it is the
only statement in a program. As shown by the examples of
Section~\ref{sec:examples}, the use of $\sleep t$ in a program does
not mean that a program necessarily waits for $t$ seconds-- depending
on the context, it may wait for anywhere between $0$ and $t$ seconds.

@@

@@

\begin{theorem}
For all programs $P$ and $Q$ then:
%%
\begin{equation}
\vtime{P} + \vtime{Q} \leq \etime{P; Q} \lesssim \etime{P} + \etime{Q}
\end{equation}
\label{theorem:main}
\end{theorem}

From these lemmas we can reason about the evaluation time of programs.
For example, consider subprograms $A$, $B$, $C$ where 
$\etime{A} = t_1$, $\etime{B} = t_2$, $\etime{C} = t_3$ and
$\vtime{A} = \vtime{B} \vtime{C} = 0$, interposed with two
sleep statements of duration $s_1$ and $s_2$:
%
\begin{equation}
\begin{array}{l}
A \\
\sleep s_1 \\
B  \\
\sleep s_2 \\
C
\end{array}
\label{example:time1}
\end{equation}
%%
Then by the above lemmas, we see that $\etime{eq. \eqref{example:time1}} = 
s_1 + s_2 + t_3$, iff $t_1 \leq 1$ and $t_2 \leq 2$.

@@

The \emph{Monad} instance for \emph{Temporal} is then as follows:
%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{37}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;\Conid{Monad}\;\Conid{Temporal}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{return}\;\Varid{a}{}\<[16]%
\>[16]{}\mathrel{=}\Conid{T}\;(\lambda \anonymous \to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{a},\Varid{vT})){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\\[-1.5em]{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}(\Conid{T}\;\Varid{p})\bind \Varid{q}{}\<[16]%
\>[16]{}\mathrel{=}\Conid{T}\;(\lambda (\Varid{startT},\Varid{nowT})\to \lambda \Varid{vT}\to {}\<[E]%
\\
\>[16]{}\hsindent{5}{}\<[21]%
\>[21]{}\mathbf{do}\;{}\<[25]%
\>[25]{}(\Varid{x},\Varid{vT'}){}\<[37]%
\>[37]{}\leftarrow \Varid{p}\;(\Varid{startT},\Varid{nowT})\;\Varid{vT}{}\<[E]%
\\
\>[25]{}\mathbf{let}\;(\Conid{T}\;\Varid{q'}){}\<[37]%
\>[37]{}\mathrel{=}\Varid{q}\;\Varid{x}{}\<[E]%
\\
\>[25]{}\Varid{thenT}{}\<[37]%
\>[37]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
\\
\>[25]{}\Varid{q'}\;(\Varid{startT},\Varid{thenT})\;\Varid{vT'}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%
To ease understanding, we recall the types of \emph{return}
and \ensuremath{(\bind )} along with some intuition for their behaviour for
\emph{Temporal}:
%
\begin{itemize}
\item \ensuremath{\Varid{return}\mathbin{::}\Varid{a}\to \Conid{Temporal}\;\Varid{a}} lifts a pure value into a trivially
effectful computation by ignoring the time parameters and 
providing the usual pure state behaviour: passing the state unchanged.

\item \ensuremath{(\bind )\mathbin{::}\Conid{Temporal}\;\Varid{a}\to (\Varid{a}\to \Conid{Temporal}\;\Varid{b})\to \Conid{Temporal}\;\Varid{b}} 
  composes two computations together.  The result of composing two
  temporal computations, with start time \emph{startT}, current time
  \emph{nowT}, and virtual time \emph{vT}, is the result of evaluating
  first the left-hand side at time \emph{nowT} and then

right hand side, and then the left hand side.  Note that
  \ensuremath{\Varid{getCurrentTime}\mathbin{::}\Conid{IO}\;\Conid{Time}} retrieves the time from the operation
  system.
\end{itemize}

The \ensuremath{\Varid{runTime}} operation executes a temporal computation
inside of the \emph{IO} monad by providing the start time of the
computation, and with virtual time 0:
%%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{runTime}\mathbin{::}\Conid{Temporal}\;\Varid{a}\to \Conid{IO}\;\Varid{a}{}\<[E]%
\\
\>[B]{}\Varid{runTime}\;(\Conid{T}\;\Varid{c})\mathrel{=}\mathbf{do}\;{}\<[21]%
\>[21]{}\Varid{startT}\leftarrow \Varid{getCurrentTime}{}\<[E]%
\\
\>[21]{}(\Varid{x},\anonymous )\leftarrow \Varid{c}\;(\Varid{startT},\Varid{startT})\;\mathrm{0}{}\<[E]%
\\
\>[21]{}\Varid{return}\;\Varid{x}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%%
To illustrate the evalution of temporal computation and the
ordering and interleaving of calls to the operation system for the
current time, we consider a program:
%%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{runTime}\;(\mathbf{do}\;\{\mskip1.5mu \Varid{f};\Varid{g};\Varid{h};\mskip1.5mu\}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
(where \ensuremath{\Varid{f}\mathrel{=}\Conid{T}\;\Varid{f'},\Varid{g}\mathrel{=}\Conid{T}\;\Varid{g'},\Varid{h}\mathrel{=}\Conid{T}\;\Varid{h'}}) which 
desugars to the following \emph{IO} computation, 
after some simplification: 
%%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{do}\;{}\<[5]%
\>[5]{}\Varid{startT}{}\<[15]%
\>[15]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
\\
\>[5]{}(\anonymous ,\Varid{vT'}){}\<[15]%
\>[15]{}\leftarrow \Varid{f'}\;(\Varid{startT},\Varid{startT})\;\mathrm{0}{}\<[E]%
\\
\>[5]{}\Varid{thenT}{}\<[15]%
\>[15]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
\\
\>[5]{}(\anonymous ,\Varid{vT''})\leftarrow \Varid{g'}\;(\Varid{startT},\Varid{thenT})\;\Varid{vT'}{}\<[E]%
\\
\>[5]{}\Varid{thenT'}{}\<[15]%
\>[15]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
\\
\>[5]{}(\Varid{y},\anonymous ){}\<[15]%
\>[15]{}\leftarrow \Varid{h'}\;(\Varid{startT},\Varid{thenT'})\;\Varid{vT''})\;{}\<[E]%
\\
\>[5]{}\Varid{return}\;\Varid{y}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%
The above illustrates the repeated calls to \ensuremath{\Varid{getCurrentTime}}, the
constant start time parameter, and the threading of virtual time state
throughout the computation.

Figure~\ref{core-functions} provides a number of effectul operations
for \emph{Temporal} that access the current time, the start time, get
and set the virtual time, and cause a kernel sleep. These
are used in the next part of the model.

@@

The key primitive \emph{sleep} provides the semantics for \sleepOp{} as:
%%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
\column{37}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{sleep}\mathbin{::}\Conid{VTime}\to \Conid{Temporal}\;(){}\<[E]%
\\
\>[B]{}\Varid{sleep}\;\Varid{delayT}\mathrel{=}\mathbf{do}\;{}\<[20]%
\>[20]{}\Varid{nowT}{}\<[30]%
\>[30]{}\leftarrow \Varid{time}{}\<[E]%
\\
\>[20]{}\Varid{vT}{}\<[30]%
\>[30]{}\leftarrow \Varid{getVirtualTime}{}\<[E]%
\\
\>[20]{}\mathbf{let}\;\Varid{vT'}{}\<[30]%
\>[30]{}\mathrel{=}\Varid{vT}\mathbin{+}\Varid{delayT}{}\<[E]%
\\
\>[20]{}\Varid{setVirtualTime}\;\Varid{vT'}{}\<[E]%
\\
\>[20]{}\Varid{startT}{}\<[30]%
\>[30]{}\leftarrow \Varid{start}{}\<[E]%
\\
\>[20]{}\mathbf{let}\;\Varid{diffT}\mathrel{=}\Varid{diffTime}\;\Varid{nowT}\;\Varid{startT}{}\<[E]%
\\
\>[20]{}\mathbf{if}\;(\Varid{vT'}\mathbin{<}\Varid{diffT})\;\mathbf{then}\;\Varid{return}\;(){}\<[E]%
\\
\>[20]{}\hsindent{17}{}\<[37]%
\>[37]{}\mathbf{else}\;\Varid{kernelSleep}\;(\Varid{vT'}\mathbin{-}\Varid{diffT}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%
Thus, \emph{sleep} proceeds by getting the current time, calculating
the new virtual time \emph{vT'} and updating the virtual time
state. If the new virtual time is less than the elapsed time
\emph{diffT} then no actual (kernel) sleeping happens. However, if the
new virtual time is ahead of the elapsed time, then the process sleeps
for the difference.

@@

\begin{proof}
By induction % possibly by strong induction?

@@

@@

For the examples of Section~\ref{sec:introduction}, the full structure
of monad is not needed to give their semantics as there is no using of
binding between statements (and thus no dataflow). In these case just
an \emph{applicative functor}~\cite{mcbride2008functional} or even a
monoid would suffice. These can be derived from the monad structure
on \emph{Temporal} since all monads are applicative functors and all 
monads define monoid over $m ()$.

@@

@@

@@

@@

\paragraph{Proof} (of Theorem~\ref{theorem:main})
Since sequential composition is associative, we can reassociate
$P; Q$ such that $P = P_1; P'$ where $P_1$ is a single statement

>>>>

The underlying programming model of SonicPi provides a way to separate
the ordering of effects from the timing of
effects. Figure~\ref{three-chord-example} shows an example program
where three chords are played in sequence, combining simple notions of 
parallel, timed, and ordered effects.

The first three statements play the notes of a C major chord in
parallel.  A \sleepOp{} statement then provides a ``temporal barrier''
which blocks the computation from continuing until 1 second has
elapsed since the \emph{start} of the program (not since the end of
playing the notes). Once one second has elapsed, the next three
statements are executed which plays an F major chord. The next
\sleepOp{} means that the final chord is not played until 1.5 seconds
has elapsed since the start of the
program. Figure~\ref{three-chord-timing} illustrates the timing.

Thus, ``$\sleep{} t$'' communicates that, after it has been evaluated, at least 
$t$ seconds has elapsed since the last \sleepOp{}. This provides a minimum
time. In between calls to \sleepOp{}, any other statements can (with some limits)
be considered task parallel.

@@

\note{Contributions}

\begin{itemize}
\item Explain the core principles of the language related to timing
(Section~\ref{}) and provide a monadic model this programming approach
(Section~\ref{}) (embedded in Haskell).

\item We show that the monadic approach can be composed with
  additional monads to capture other useful notions of effect in
  \lang{} programs, such as random numbers (Section~\ref{}).
\end{itemize}

%
The \lang{} was previously discussed in~\cite{aaron2013sonic}

@@

Figure~\ref{sleep-examples} shows four similar programs which each
have different internal behaviours for \sleepOp, illustrating its semantics. 
The first three take 3s to execute and the
last takes 4s to execute, with the behaviours:
%
\begin{enumerate}[(a)]
\item{3s -- sleeps for 1s then sleeps for 2s (two sleeps performed);}
\item{3s -- performs a computation lasting 1s, ignores
the first \sleepOp{} since its minimum duration has been reached, 
and then sleeps for 2s (one sleep performed);}
\item{3s -- performs a computation lasting 2s, which means that
the first \sleepOp{} is ignored, and the second \sleepOp{} waits
for only 1s to reach its minimum duration (half a sleep performed);}
\item{4s -- performs a computation lasting 2s, thus 
the first \sleepOp{} is ignored, then performs a computation lasting
2s, thus the second \sleepOp{} is ignored (no sleeps performed).}
\end{enumerate}

@@

The \sleepOp{} operation is the way to change the virtual
time. Therefore, the definition of $\vtime{-}$ can be easily defined
over all programs:
%
\begin{definition}
Virtual time is specified for statements of \lang{} programs 
by the following (ordered) cases:
%
\begin{align*}
\vtime{\sleep t} & = t \\ 
\vtime{P; Q} & = \vtime{P} + \vtime{Q} \\
\vtime{-} & = 0
\end{align*}
%
\ie{}, the virtual time is $0$ 
for any statment other than \sleepOp{} or sequential composition.
\label{sleep-spec}
\end{definition}
\note{I haven't included here expressions, for example, if a call
to a function that did some sleeping happened, then this is not accounted
for here. I could be easily incldued though.}

\paragraph{Equality on time}

Providing exact deadlines in real-time systems is difficult due 
to noise and overhead caused by execution. We do not ignore
this problem in the programming model of \lang{} and the discussion here. 
We define the relation $\approx$ on actual time (non virtual times), where:
%%
\begin{equation}
s \approx t
\;
\equiv
\;
(t - \epsilon) \leq s \leq (t + \epsilon)
\end{equation}
%
for some value of $\epsilon$ which is the maximum negligible 
time value with respect to the application at hand. In the case of
\lang{}, this is equal to the scheduling delay for the SuperCollider
engine (and is machine dependent), which is roughly \note{X}.

@@

The virtual time and actual time of a single sleep statement 
 are roughly the same, \ie{}, $\vtime{\sleep t} \approx
\etime{\sleep t}$ and thus $\etime{\sleep t} \approx t$ (by the
specification in Definition~\ref{sleep-spec}). This holds
only when \sleepOp{} is used in isolation, that is, when it is the
only statement in a program. As shown by the examples of
Section~\ref{sec:examples}, the use of $\sleep t$ in a program does
not mean that a program necessarily waits for $t$ seconds-- depending
on the context, it may wait for anywhere between $0$ and $t$ seconds.

\paragraph{Temporal properties of programs}

We outline here some important temporal properties of our \lang{} programs
that relates the virtual time and actual times. In Section~\ref{}, we
replay these lemmas and prove the soundness: that these lemmas hold of our model.

@@

\noindent
These two lemmas illuminate something of the semantics of sleep,
and its interaction with other statements in the language.

@@

\begin{lemma}
For all programs $P$ and $Q$ then:
%%
\begin{equation}
\vtime{P} + \vtime{Q} \leq \etime{P; Q} \lesssim \etime{P} + \etime{Q}
\end{equation}
\label{theorem:main}
\end{lemma}

\note{It's possible that these lemmas should be in a different order--
 we may need \ref{theorem:main} to prove \ref{lemma-rel-etime-vtime+}.}

From these lemmas we can reason about the evaluation time of 
programs. For example, consider subprograms $A$, $B$, $C$ where 
$\vtime{A} = \vtime{B} = \vtime{C} = 0$ interposed with two
sleep statements of duration $s_1$ and $s_2$:
%
\begin{equation}
\begin{array}{l}
A \\
\sleep s_1 \\
B  \\
\sleep s_2 \\
C
\end{array}
\label{example:time1}
\end{equation}
%%
Then by the above lemmas, we see that if $\etime{A} \leq s_1$ and
$\etime{B} \leq s_2$ then $\etime{eq. \eqref{example:time1}} = 
s_1 + s_2 + \etime{C}$.

@@

The \emph{Monad} instance for \emph{Temporal} is then as follows:
%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{37}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;\Conid{Monad}\;\Conid{Temporal}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{return}\;\Varid{a}{}\<[16]%
\>[16]{}\mathrel{=}\Conid{T}\;(\lambda \anonymous \to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{a},\Varid{vT})){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\\[-1.5em]{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}(\Conid{T}\;\Varid{p})\bind \Varid{q}{}\<[16]%
\>[16]{}\mathrel{=}\Conid{T}\;(\lambda (\Varid{startT},\Varid{nowT})\to \lambda \Varid{vT}\to {}\<[E]%
\\
\>[16]{}\hsindent{5}{}\<[21]%
\>[21]{}\mathbf{do}\;{}\<[25]%
\>[25]{}(\Varid{x},\Varid{vT'}){}\<[37]%
\>[37]{}\leftarrow \Varid{p}\;(\Varid{startT},\Varid{nowT})\;\Varid{vT}{}\<[E]%
\\
\>[25]{}\mathbf{let}\;(\Conid{T}\;\Varid{q'}){}\<[37]%
\>[37]{}\mathrel{=}\Varid{q}\;\Varid{x}{}\<[E]%
\\
\>[25]{}\Varid{thenT}{}\<[37]%
\>[37]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
\\
\>[25]{}\Varid{q'}\;(\Varid{startT},\Varid{thenT})\;\Varid{vT'}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%
To ease understanding, we recall the types of \emph{return}
and \ensuremath{(\bind )} along with some intuition for their behaviour for
\emph{Temporal}:
%
\begin{itemize}
\item \ensuremath{\Varid{return}\mathbin{::}\Varid{a}\to \Conid{Temporal}\;\Varid{a}} lifts a pure value into a trivially
effectful computation by ignoring the time parameters and 
providing the usual pure state behaviour of returning the parameter state unchanged
(named \emph{vT} in this case).

\item \ensuremath{(\bind )\mathbin{::}\Conid{Temporal}\;\Varid{a}\to (\Varid{a}\to \Conid{Temporal}\;\Varid{b})\to \Conid{Temporal}\;\Varid{b}} 
  composes two computations together.  The result of composing two
  temporal computations, with start time \emph{startT}, current time
  \emph{nowT}, and virtual time \emph{vT}, is the result of evaluating
  first the left-hand side at time \emph{nowT} and then right-hand side 
  at the new current time \emph{thenT}.

  The expression \ensuremath{\Varid{getCurrentTime}\mathbin{::}\Conid{IO}\;\Conid{Time}} retrieves the time from
  the operation system.
\end{itemize}

To model the evaluation of a program, the \ensuremath{\Varid{runTime}} operation executes
a temporal computation inside of the \emph{IO} monad by providing the
start time of the computation and virtual time 0:
%%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{runTime}\mathbin{::}\Conid{Temporal}\;\Varid{a}\to \Conid{IO}\;\Varid{a}{}\<[E]%
\\
\>[B]{}\Varid{runTime}\;(\Conid{T}\;\Varid{c})\mathrel{=}\mathbf{do}\;{}\<[21]%
\>[21]{}\Varid{startT}\leftarrow \Varid{getCurrentTime}{}\<[E]%
\\
\>[21]{}(\Varid{x},\anonymous )\leftarrow \Varid{c}\;(\Varid{startT},\Varid{startT})\;\mathrm{0}{}\<[E]%
\\
\>[21]{}\Varid{return}\;\Varid{x}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%%
To illustrate the evalution of temporal computation and the
ordering and interleaving of calls to the operation system for the
current time, consider the program:
%%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{runTime}\;(\mathbf{do}\;\{\mskip1.5mu \Varid{f};\Varid{g};\Varid{h};\mskip1.5mu\}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
(where \ensuremath{\Varid{f}\mathrel{=}(\Conid{T}\;\Varid{f'}),\Varid{g}\mathrel{=}(\Conid{T}\;\Varid{g'}),\Varid{h}\mathrel{=}(\Conid{T}\;\Varid{h'})}) which 
desugars to the following \emph{IO} computation, 
after some simplification: 
%%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{do}\;{}\<[5]%
\>[5]{}\Varid{startT}{}\<[15]%
\>[15]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
\\
\>[5]{}(\anonymous ,\Varid{vT'}){}\<[15]%
\>[15]{}\leftarrow \Varid{f'}\;(\Varid{startT},\Varid{startT})\;\mathrm{0}{}\<[E]%
\\
\>[5]{}\Varid{thenT}{}\<[15]%
\>[15]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
\\
\>[5]{}(\anonymous ,\Varid{vT''})\leftarrow \Varid{g'}\;(\Varid{startT},\Varid{thenT})\;\Varid{vT'}{}\<[E]%
\\
\>[5]{}\Varid{thenT'}{}\<[15]%
\>[15]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
\\
\>[5]{}(\Varid{y},\anonymous ){}\<[15]%
\>[15]{}\leftarrow \Varid{h'}\;(\Varid{startT},\Varid{thenT'})\;\Varid{vT''})\;{}\<[E]%
\\
\>[5]{}\Varid{return}\;\Varid{y}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%
This illustrates the repeated calls to \ensuremath{\Varid{getCurrentTime}}, the
constant start time parameter, and the threading of virtual time state
throughout the computation.

Figure~\ref{core-functions} shows a number of effectul operations of 
 the \emph{Temporal} monad that access the current time, the start time, get
and set the virtual time, and cause a kernel sleep. These
are used in the next part of the model.

@@

The key primitive \emph{sleep} provides the semantics for \sleepOp{} as:
%%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{sleep}\mathbin{::}\Conid{VTime}\to \Conid{Temporal}\;(){}\<[E]%
\\
\>[B]{}\Varid{sleep}\;\Varid{delayT}\mathrel{=}\mathbf{do}\;{}\<[20]%
\>[20]{}\Varid{nowT}{}\<[30]%
\>[30]{}\leftarrow \Varid{time}{}\<[E]%
\\
\>[20]{}\Varid{vT}{}\<[30]%
\>[30]{}\leftarrow \Varid{getVirtualTime}{}\<[E]%
\\
\>[20]{}\mathbf{let}\;\Varid{vT'}{}\<[30]%
\>[30]{}\mathrel{=}\Varid{vT}\mathbin{+}\Varid{delayT}{}\<[E]%
\\
\>[20]{}\Varid{setVirtualTime}\;\Varid{vT'}{}\<[E]%
\\
\>[20]{}\Varid{startT}{}\<[30]%
\>[30]{}\leftarrow \Varid{start}{}\<[E]%
\\
\>[20]{}\mathbf{let}\;\Varid{diffT}\mathrel{=}\Varid{diffTime}\;\Varid{nowT}\;\Varid{startT}{}\<[E]%
\\
\>[20]{}\mathbf{if}\;(\Varid{vT'}\mathbin{<}\Varid{diffT}){}\<[E]%
\\
\>[20]{}\hsindent{2}{}\<[22]%
\>[22]{}\mathbf{then}\;\Varid{return}\;(){}\<[E]%
\\
\>[20]{}\hsindent{2}{}\<[22]%
\>[22]{}\mathbf{else}\;\Varid{kernelSleep}\;(\Varid{vT'}\mathbin{-}\Varid{diffT}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%
where \emph{sleep} proceeds first by getting the current time
\emph{nowT}, calculating the new virtual time \emph{vT'} and updating
the virtual time state. If the new virtual time is less than the
elapsed time \emph{diffT} since the start then no actual (kernel)
sleeping happens. However, if the new virtual time is ahead of the
elapsed time, then the process waits for the difference such that the
elapsed time equals the virtual time.

@@

\note{Rough notes for the proof}
\begin{proof}
By induction % possibly by strong induction?

@@

\note{This is actually quite hard. I think I can do it
by using Lemma 1 and 2 together with some reassociating, but
haven't time to sort it right now.}

@@

For the examples of Section~\ref{sec:introduction}, the full structure
of monad is not needed to give their semantics as there is no using of
binding between statements (and thus no dataflow). In these case just
an \emph{applicative functor}~\cite{mcbride2008functional} or even a
monoid would suffice. These can be derived from the monad structure
on \emph{Temporal} since all monads are applicative functors and all 
monads $m$ define a monoid over \ensuremath{\Varid{m}\;()}.

@@

\note{TODO: report warnings when overrun occurs- should be easy I think
by adding stuff to sleep.}

@@

\note{TOOD: Again pretty easy, just need some info in the monad
to say what counts as a bad overrun}

@@

\note{TODO}

@@

\paragraph{Proof} (of Theorem~\ref{theorem:main})
\note{These are just rough notes}
Since sequential composition is associative, we can reassociate
$P; Q$ such that $P = P_1; P'$ where $P_1$ is a single statement
